Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-06-08 Manuscript/MPESubmission/TEM_draft.tex
Reformatted to MPE format

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

\usepackage{changepage}
\usepackage{color}
\usepackage[utf8]{inputenc}
\usepackage{textcomp,marvosym}
\usepackage{fixltx2e}
\usepackage{amsmath,amssymb}
\usepackage{cite}
\usepackage{nameref,hyperref}
\usepackage[right]{lineno}
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
\usepackage{rotating}
\usepackage{setspace} 
\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

\bibliographystyle{plos2009}

\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[natwidth=1.3in,natheight=0.4in]{PLOSlogo.png}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf\newline{Effects of missing data on topological inference using a Total Evidence approach}
}
\newline
\\
Thomas Guillerme\textsuperscript{1,2,*},
Natalie Cooper\textsuperscript{1,2}
\\
\bf{1} School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland
\\
\bf{2} Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland
\\
* E-mail: Corresponding guillert@tcd.ie
\end{flushleft}
\section*{Abstract}
To fully understand macroevolutionary patterns and processes, we need to include both extant and extinct species in our models. This requires phylogenetic trees with both living and fossil taxa at the tips. One way to infer such phylogenies is the Total Evidence approach which uses molecular data from living taxa and morphological data from living and fossil taxa.

@@

\linenumbers

@@

\section*{Introduction}
Although most species that have ever lived are now extinct \cite{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species (e.g. \cite{meredithimpacts2011,jetzthe2012}). Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events (e.g. \cite{pyrondivergence2011}), relationships among lineages (e.g. \cite{manosphylogeny2007}) or niche occupancy (e.g. \cite{pearmanniche2008}). This has led to increasing consensus among evolutionary biologists that fossil taxa should be included in macroevolutionary studies \cite{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments (e.g. \cite{pyrondivergence2011,ronquista2012,BEASTmaster}).

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies. These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny, and in which part of the available fossil data is used (i.e. the age of the fossil only or both its age and morphology). Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \cite{simpson1945}. This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty. Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species. Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these phylogenies and their occurrence dates are used to time calibrate phylogenies \cite{zuckerkandl1965}. There have been great improvements in the theory and application of these two approaches (e.g. \cite{bapsta2013,stadlerdating2013,heaththe2013}) as well as much debate about the ``best" approach to use (e.g. \cite{spencerefficacy2013,wrightbayesian2014}). However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \cite{eernissetaxonomic1993}. This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny (known as tip-dating \cite{ronquista2012}), and allows the use of probabilistic methods for estimating phylogenetic uncertainty \cite{ronquista2012}. Total Evidence methods have been successfully applied to empirical data (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}), and are becoming an increasingly popular way of adding fossil taxa to phylogenies.
However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires both molecular and morphological data, both of which can be difficult (or impossible) to collect for every living and fossil taxon in the tree.
Morphological data for living taxa are rarely collected when molecular data are available (e.g. \cite{O'Leary08022013} \textit{vs.} \cite{meredithimpacts2011}), and for fossil taxa, data can only be collected from features preserved in the fossil record (for example, in vertebrates, the hardest parts of the skeleton are more often preserved than soft parts \cite{sansomfossilization2013}; and molecular data is (nearly) always unavailable. Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \cite{salamin2003}.

Although missing data does not appear be a major problem in molecular and morphological matrices separately (as long as enough data overlaps in each case, and missing data is not biased towards particular subclades \cite{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}), it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa. This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices. Until now, no attempt has been made to study the impact of this issue on phylogenetic inference using Total Evidence methods.

In this study, we focus on the effect of missing data on our ability to recover a ``best" tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species (e.g. \cite{meredithimpacts2011,jetzthe2012}), or for studying evolutionary transitions (e.g. \cite{friedmanexplosive2010}). Although branch length estimation is also important (namely for timing extinction and/or speciation events - e.g. \cite{ronquista2012}), we do not consider branch lengths in this study. This is partially due to difficulties with simulating branch lengths and topology simultaneously, but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but with a fixed topology approach \cite{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}. Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \cite{ronquista2012}, the effect of missing data on such matrices is well known \cite{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}. Therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix). We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best" tree topology more so than minimizing the amount of missing data in the fossil record. Additionally, we find that the ability of Total Evidence methods to recover the ``best" tree topology is increased when using Bayesian methods.

@@

\section*{Methods}
To explore how missing data in the morphological sections of Total Evidence matrices influences tree topology, we used the following protocol (note that we explain each step in detail below this general outline; Fig.~\ref{Fig_Outline}).
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true" tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete" matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). We call the resulting 125 matrices ``missing-data" matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete" matrix and from the 125 ``missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the ``best" tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data" trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best" tree to the ``missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

\begin{figure}[h]
\caption{{\bf Protocol outline.}
(1) We randomly generated a birth-death tree (the ``true" tree) and used it to simulate a matrix with no missing data (the ``complete" matrix).
(2) We removed data from the morphological part of the ``complete" matrix resulting in 125 ``missing-data" matrices.
(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
(4) We compared the ``missing-data" trees to the ``best" tree.
We repeated these four steps 50 times.}
\label{Fig_Outline}
\end{figure}

@@

\subsection*{1. Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a ``true" tree of 50 taxa in R v. 3.0.2 \cite{R302} using the package diversitree v. 0.9-6 \cite{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution (bounded between 0 and 1) but maintaining $\lambda$ $>$ $\mu$ \cite{paradistime-dependent2011}. Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa. For example, fossil taxa make up 88\% \cite{beckancient2014}, 58\% \cite{schragocombining2013}, 48\% \cite{pyrondivergence2011}, 31\% \cite{ronquista2012} and 31\% \cite{slaterphylogenetic2013} of taxa in various studies. To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa. The fossil taxa were considered as unique tips at the end of extinct lineages. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we generated a molecular and a morphological matrix from the ``true" tree. The molecular matrix was inferred from the ``true" tree using the R package phyclust v. 0.1-14 \cite{chen2011}. The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \cite{ranbaut1997seqgen} and using the HKY model \cite{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \cite{douadycomparison2003}. The substitution rates were distributed following a gamma distribution with an alpha ($\alpha$) shape of 0.5 \cite{yangamong-site1996}. We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years). All the molecular information for fossil taxa was replaced by missing data ("?").

We inferred the morphological matrix using the R package ape v. 3.0-11 \cite{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (see ~\nameref{S1_Appendix1} and Fig. S1 within). We then ran an independent discrete character simulation for each character using the ``true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \cite{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \cite{wagner2000,davalosintegrating2014,wrightbayesian2014}.

@@

\subsection*{2. Removing data}
\label{Removing_data}
We modified the ``complete" matrix to get matrices with missing data by randomly replacing data with ``?" in the morphological part of the matrices according to the following parameters:
\begin{enumerate}
\item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that the increased availability of molecular data means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
\item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record, a common problem due to preservation biases \cite{sansomfossilization2013}. 
\item{$M_{C}$, the proportion of missing morphological characters for both living and fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa. This reflects the amount of effort put into identifying the characters (e.g. \cite{O'Leary08022013}).
\end{enumerate}

@@

\subsection*{3. Estimating phylogenies}
From the resulting matrices we generated two types of trees: the ``best" tree inferred from the ``complete" matrix and the ``missing-data" trees inferred from the 125 matrices with various amounts of missing data. The ``true" tree was used to generate the ``complete" matrix and reflects the ``true" evolutionary history in our simulations. The ``best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods. In real world situations, the ``true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \cite{rozen2005}). Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the ``best" tree, rather than the ``true" tree, as the ``best" tree is generally what we have to work with.

\paragraph*{Maximum Likelihood}
The ``best" tree and the ``missing-data" trees were inferred using RAxML v. 8.0.20 \cite{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model \cite{tavare1986} (default GTRGAMMA in RAxML v. 8.0.20  \cite{Stamatakis21012014}). For the morphological data, we used the implemented M\textit{kv} model \cite{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20 \cite{Stamatakis21012014}). We used RAxML because it automatically corrects for acquisition bias \cite{lewisa2001}. It is also heavily used in the literature for Maximum Likelihood tree inference (e.g. \cite{rouresite-specific2011,Bogdanowicz2012,springermacroevolutionary2012,O'Leary08022013,kellymolecular2014}) and is one of the fastest methods available \cite{Stamatakis01102008}.

>>>>

\documentclass[12pt,letterpaper]{article}

%Packages
\usepackage{xcolor}
\usepackage{changepage}
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{natbib}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{longtable}
\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
%\setcounter{secnumdepth}{0}

@@

%Running head
\begin{flushright}
%Version dated: \today
\end{flushright}
\bigskip
\noindent Running head: Missing data and topology in Total Evidence method

\bigskip
\medskip
\begin{center}

\noindent{\Large \bf Effects of missing data on topological inference using a Total Evidence approach}

\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^a$$^,$$^b$$^*$, and Natalie Cooper$^a$$^,$$^b$$^,$$^1$}\\
\noindent {\small \it 
$^a$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
$^b$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\
$^*$Corresponding author. Zoology Building, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.\\
$^1$Present address: Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK. E-mail: nhcooper12@gmail.com}\\
\end{center}

\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

\newpage
\section* {Abstract}
To fully understand macroevolutionary patterns and processes, we need to include both extant and extinct species in our models. This requires phylogenetic trees with both living and fossil taxa at the tips. One way to infer such phylogenies is the Total Evidence approach which uses molecular data from living taxa and morphological data from living and fossil taxa.

@@

\bigskip
\noindent
\textbf{Keywords:} morphological characters, Bayesian, Maximum Likelihood, topology, fossil, living.

@@

\newpage
\section{Introduction}
Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species (e.g. \citealp{meredithimpacts2011,jetzthe2012}). Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages (e.g. \citealp{manosphylogeny2007}) or niche occupancy (e.g. \citealp{pearmanniche2008}). This has led to increasing consensus among evolutionary biologists that fossil taxa should be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments (e.g. \citealp{pyrondivergence2011,ronquista2012,BEASTmaster}).

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies. These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny, and in which part of the available fossil data is used (i.e. the age of the fossil only or both its age and morphology). Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{simpson1945}. This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty. Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species. Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these phylogenies and their occurrence dates are used to time calibrate phylogenies \citep{zuckerkandl1965}. There have been great improvements in the theory and application of these two approaches (e.g. \citealp{bapsta2013,stadlerdating2013,heaththe2013}) as well as much debate about the ``best" approach to use (e.g. \citealp{spencerefficacy2013,wrightbayesian2014}). However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}. This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny (known as tip-dating \citealp{ronquista2012}), and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}. Total Evidence methods have been successfully applied to empirical data (e.g. \citealp{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}), and are becoming an increasingly popular way of adding fossil taxa to phylogenies.
However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires both molecular and morphological data, both of which can be difficult (or impossible) to collect for every living and fossil taxon in the tree.
Morphological data for living taxa are rarely collected when molecular data are available (e.g. \citealp{O'Leary08022013} \textit{vs.} \citealp{meredithimpacts2011}), and for fossil taxa, data can only be collected from features preserved in the fossil record (for example, in vertebrates, the hardest parts of the skeleton are more often preserved than soft parts \citep{sansomfossilization2013}; and molecular data is (nearly) always unavailable. Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.

Although missing data does not appear be a major problem in molecular and morphological matrices separately (as long as enough data overlaps in each case, and missing data is not biased towards particular subclades \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa. This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices. Until now, no attempt has been made to study the impact of this issue on phylogenetic inference using Total Evidence methods.

In this study, we focus on the effect of missing data on our ability to recover a ``best" tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species (e.g. \citealp{meredithimpacts2011,jetzthe2012}), or for studying evolutionary transitions (e.g. \citealp{friedmanexplosive2010}). Although branch length estimation is also important (namely for timing extinction and/or speciation events - e.g. \citealp{ronquista2012}), we do not consider branch lengths in this study. This is partially due to difficulties with simulating branch lengths and topology simultaneously, but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but with a fixed topology approach \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}. Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}. Therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix). We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best" tree topology more so than minimizing the amount of missing data in the fossil record. Additionally, we find that the ability of Total Evidence methods to recover the ``best" tree topology is increased when using Bayesian methods.

@@

\section{Materials and Methods}
To explore how missing data in the morphological sections of Total Evidence matrices influences tree topology, we used the following protocol (note that we explain each step in detail below this general outline; Figure 1).
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true" tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete" matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). We call the resulting 125 matrices ``missing-data" matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete" matrix and from the 125 ``missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the ``best" tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data" trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best" tree to the ``missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

%\begin{figure}[h]
%\caption{{\bf Protocol outline.}
%(1) We randomly generated a birth-death tree (the ``true" tree) and used it to simulate a matrix with no missing data (the ``complete" matrix).
%(2) We removed data from the morphological part of the ``complete" matrix resulting in 125 ``missing-data" matrices.
%(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
%(4) We compared the ``missing-data" trees to the ``best" tree.
%We repeated these four steps 50 times.}
%\label{Fig_Outline}
%\end{figure}

@@

\subsection{Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a ``true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution (bounded between 0 and 1) but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}. Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa. For example, fossil taxa make up 88\% \citep{beckancient2014}, 58\% \citep{schragocombining2013}, 48\% \citep{pyrondivergence2011}, 31\% \citep{ronquista2012} and 31\% \citep{slaterphylogenetic2013} of taxa in various studies. To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa. The fossil taxa were considered as unique tips at the end of extinct lineages. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we generated a molecular and a morphological matrix from the ``true" tree. The molecular matrix was inferred from the ``true" tree using the R package phyclust v. 0.1-14 \citep{chen2011}. The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \citep{douadycomparison2003}. The substitution rates were distributed following a gamma distribution with an alpha ($\alpha$) shape of 0.5 \citep{yangamong-site1996}. We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years). All the molecular information for fossil taxa was replaced by missing data ("?").

We inferred the morphological matrix using the R package ape v. 3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (see Appendix A and Fig. A1 within). We then ran an independent discrete character simulation for each character using the ``true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014,wrightbayesian2014}.

@@

\subsection{Removing data}
\label{Removing_data}
We modified the ``complete" matrix to get matrices with missing data by randomly replacing data with ``?" in the morphological part of the matrices according to the following parameters:
\begin{enumerate}
\item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that the increased availability of molecular data means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
\item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record, a common problem due to preservation biases \citep{sansomfossilization2013}. 
\item{$M_{C}$, the proportion of missing morphological characters for both living and fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa. This reflects the amount of effort put into identifying the characters (e.g. \citealp{O'Leary08022013}).
\end{enumerate}

@@

\subsection{Estimating phylogenies}
From the resulting matrices we generated two types of trees: the ``best" tree inferred from the ``complete" matrix and the ``missing-data" trees inferred from the 125 matrices with various amounts of missing data. The ``true" tree was used to generate the ``complete" matrix and reflects the ``true" evolutionary history in our simulations. The ``best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods. In real world situations, the ``true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \citealp{rozen2005}). Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the ``best" tree, rather than the ``true" tree, as the ``best" tree is generally what we have to work with.

\subsubsection{Maximum Likelihood}
The ``best" tree and the ``missing-data" trees were inferred using RAxML v. 8.0.20 \citep{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model \citep{tavare1986} (default GTRGAMMA in RAxML v. 8.0.20  \citep{Stamatakis21012014}). For the morphological data, we used the implemented M\textit{kv} model \citep{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20 \citep{Stamatakis21012014}). We used RAxML because it automatically corrects for acquisition bias \citep{lewisa2001}. It is also heavily used in the literature for Maximum Likelihood tree inference (e.g. \citealp{rouresite-specific2011,Bogdanowicz2012,springermacroevolutionary2012,O'Leary08022013,kellymolecular2014}) and is one of the fastest methods available \citep{Stamatakis01102008}.

To measure the support for each branch in our simulated phylogenies we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the ``complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete" and 50 x 125 ``missing-data" matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50). This step was implemented to make sure that the differences we observed in topologies (see below) were due to the amount of missing data for each parameter ($M_L$, $M_F$ and $M_C$) and not simply to low branch support that is likely to lead to different topologies. On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}. Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes). We performed this procedure to increase the resolution of our resulting trees.
